---
title: "HAL-QuIC"
subtitle: "Kinetic Modeling of the RT-QuIC Assay"
author: "Gage Rowden"
date: today
format:
  revealjs:
    theme: 
        - static/custom.scss
    slide-number: true
    transition: fade
    chalkboard: false
    mermaid:
      theme: dark
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r Initialize}
#| include: false
library(tidyverse)
library(quicR)
library(janitor)
library(zoo)
library(modelr)
library(investr)
library(magrittr)
library(latex2exp)
library(ggpubr)

font_title <- "Montserrat"
font_text <- "Montserrat"
font_sans <- "Noto Sans" 
color_text <- "#EEEEEE"
color_base <- "#252525"
color_base_light <- "#95DFD6"
color_bg <- "#151515"
color_accent <- "#B75C9D"
color_accent_light <- "#DBA6CC"
gray <- "#bebebe"
darkgray <- "#6d6d6d"
primary_color <- "#ff4f4f"
secondary_color <- "#54a4ff"


list.files("R", full.names = TRUE) %>%
  walk(source)

main_theme <- theme(
    plot.background = element_rect(fill = color_bg),
    panel.grid = element_line(color = color_base),
    panel.background = element_rect(fill = color_bg),
    axis.line.y  = element_line(color = primary_color),
    axis.ticks.y = element_line(color = primary_color),
    axis.text.y  = element_text(color = primary_color),
    axis.title.y = element_text(color = primary_color),
    axis.line.y.right  = element_line(color = secondary_color),
    axis.ticks.y.right = element_line(color = secondary_color),
    axis.text.y.right  = element_text(color = secondary_color),
    axis.title.y.right = element_text(color = secondary_color)
)
```

# Background

## What is RT-QuIC?

Real-Time Quaking-Induced Conversion is a seed amplification assay (SAA) used to detect misfolded prion protein (PrP^Sc^) [@atarashi2011].

- A small amount of seed material (brain or lymph node homogenate) is incubated with recombinant substrate protein (PrP^C^) under cyclic shaking
- Prion seeds template misfolding of the substrate, forming amyloid aggregates
- Aggregation is tracked in real time via thioflavin T (ThT) fluorescence
- The result is a **sigmoidal fluorescence trace**, characteristic of nucleated self-assembly [@ferrone1999]

## Prion Fibril Kinetics

A sigmoid would normally suggest cooperative binding — but PrP^Sc^ is a linear fibril, not an allosteric complex.

@masel1999 proposed a **nucleated polymerization** model instead, tracking monomer, fibril number, and aggregated mass:

$$
\frac{dM}{dt}=-nk_{n}M^{n}-k_{g}MP \qquad
\frac{dP}{dt}=k_{n}M^{n}+k_{f}Z-k_{f}M_{0}P
$$

Because RT-QuIC seeds with a **mature** fibril, spontaneous nucleation ($k_n$) becomes negligible, simplifying to:

$$
\frac{dP}{dt}=k_fZ \qquad\qquad \frac{dZ}{dt}=k_gMP
$$

## Measuring the Reaction

ThT fluorescence tracks fibril formation, but isn't a direct measure of substrate consumption. @lee2007 paired it with UV-vis absorbance to track both sides of the reaction simultaneously.

```{r Absorbance vs ThT}
#| fig-height: 5
#| fig-width: 9
#| fig-align: center
a <- 0.25
b <- 36

tibble(
  time = seq(0, 72, 0.25),
  tht  = 1 / (1 + exp(a * (b - time))),
  abs  = 1 / (1 + exp(a * (time - b)))
) %>%
  pivot_longer(c(tht, abs), names_to = "measure", values_to = "value") %>%
  ggplot(aes(time, value, color = measure)) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(sec.axis = sec_axis(~ .x, name = "Fluorescence (Fibril %)")) +
  scale_color_manual(values = c(primary_color, secondary_color)) +
  labs(x = "Time (hr)", y = "Absorbance (Substrate %)") +
  main_theme +
  theme(
    legend.position = "none",
  )
```

## A Missing Piece

Standard RT-QuIC analysis reports one number: the **lag time to threshold** (rate of amyloid formation, RAF).

But normalized fluorescence traces frequently show more structure:

- A rapid **growth** phase, then
- A **dampening** phase that pulls the signal back down after the peak

This second phase is poorly captured by a single sigmoid — and it's currently ignored by standard analysis.

# How the Model Is Generated

## The Double Sigmoidal Model

HAL-QuIC models the full trace as the **sum of two logistic functions**:

$$
f(t)=\frac{S_1}{1+e^{a_1(b_1-t)}}+\frac{S_2}{1+e^{a_2(b_2-t)}}
$$

```{r Example Graph}
#| fig-height: 5
#| fig-width: 9
example_graph(S1 <- 25, S2 = -12.5, a1 = 1.5, a2 = 0.5, b1 = 10, b2 = 16)
```

## Parameter Interpretation

Each sigmoid contributes three parameters — six in total describe a reaction:

| Parameter | Meaning |
|:---:|---|
| $S_1$, $S_2$ | Asymptotes of 1$^\circ$ / 2$^\circ$ phases |
| $a_1$, $a_2$ | Steepness of inflection points |
| $b_1$, $b_2$ | Time to inflection point |

Together, $(S_1, a_1, b_1, S_2, a_2, b_2)$ form a compact **kinetic fingerprint** for a single RT-QuIC reaction.

## Data Processing Pipeline

```{mermaid}
flowchart LR
  A["Raw fluorescence\n(.parquet)"] --> B["normalize()\nsmooth + normalize + derivatives"]
  B --> C["calculate_metrics()\nMPR, AUC, time-to-threshold"]
  C --> D["estimate_params()\nlandmark starting values"]
  D --> E["fit_model()\ndouble sigmoidal nls()"]
  E --> F["results.parquet\ncoefficients + derived params"]
```

Landmark values (peak fluorescence, inflection times, decay characteristics) are extracted directly from the smoothed data and its derivatives, giving the nonlinear fitter informed starting values instead of arbitrary guesses.

## Fitting Procedure

`fit_model()` uses bounded `nls()` with `algorithm = "port"`, and a 10-second timeout per call:

1. Attempt the **double sigmoidal** fit using landmark-derived starting values
2. If it fails, fall back to a **single sigmoidal** fit
3. Re-attempt the **double** fit, seeded with the single fit's coefficients
4. If that also fails, report the single sigmoidal result

This staged fallback keeps the pipeline resilient across thousands of individual well reactions with widely varying signal quality.

## Derived Parameters

From the six fitted coefficients, three additional kinetic quantities are calculated:

| Parameter | Formula | Meaning |
|:---:|:---:|---|
| $K_{app}$ | $1 / b_1$ | Apparent rate constant |
| $t_{lag}$ | $b_1 - \dfrac{S_1}{2a_1}$ | Lag time (x-intercept of tangent at inflection) |
| $t_{lag}'$ | $b_1 - \dfrac{a_1}{2}$ | Alternative lag time estimate |

These give a bridge back to conventional RT-QuIC metrics (like RAF) while retaining the full shape information the double sigmoid captures.

# Current Work: Describing Reaction Kinetics

## Research Roadmap

The open question: is the secondary phase a genuine kinetic phenomenon, or an assay artifact? Experiments are tiered by required effort.

```{mermaid}
flowchart LR
  subgraph T1["Tier 1 — Assay Parameters"]
    direction TB
    E1(["Seed Concentration"])
    E2(["Fragmentation Dynamics"])
    E7(["Substrate Concentration"])
    E1-->E7
    E7-->E2
  end

  subgraph T2["Tier 2 — Measurement Validation"]
    direction TB
    E3(["Absorbance & Fluorescence"])
    E4(["Fibril/Soluble Protein"])
    E3-->E4
  end

  subgraph T3["Tier 3 — Generalizability"]
    direction TB
    E5(["Prion Strain Effects"])
    E6(["Substrate Species Effects"])
    E5-->E6
  end

  T1 --> T2 --> T3
```

## Experiment 1 — Seed Concentration {.smaller}

**Question:** Does seed concentration modulate the amplitude, timing, or inflection of the secondary phase?

**Design:** 10-fold serial dilution of a known CWD-positive lymph node sample, run under standard conditions (42°C, 700 rpm, 15 min read cycle, 96 hr).

```{r Load seed data}
#| include: false
seed_files <- list.files("raw/seed_effects", pattern = "*.xlsx", full.names = TRUE)
seed_groups <- c("sample", "well", "dilutions", "rxn")

get_raw <- function(file) {
  file %>%
    get_quic() %>%
    mutate(
      rxn = str_extract(file, "2026\\d{4}_r\\d+_[A-Z]+_\\D+\\."),
      rxn = str_remove(rxn, "\\.")
    )
}

seed_raw <- seed_files %>%
  map_dfr(get_raw) %>%
  clean_names(replace = c("Sample IDs" = "sample")) %>%
  filter(time > 0) %>%
  normalize("time", "rfu", 8, seed_groups, 3) %>%
  mutate(
    across(c(sample, well, rxn), as.factor),
    dilutions = -log10(as.integer(dilutions))
  ) %>%
  na.omit()

seed_df <- seed_raw %>%
  nest(.by = all_of(seed_groups)) %>%
  left_join(
    calculate_metrics(
      seed_raw, seed_groups, threshold = 3,
      time_col = "time", ttt_values = "norm", auc_values = "norm",
      norm_col = "norm", deriv_col = "deriv"
    ),
    by = seed_groups
  ) %>%
  filter(MPR > 3) %>%
  estimate_params() %>%
  na.omit()

seed_mod <- seed_df %>%
  mutate(model = pmap(., fit_model, single_only = FALSE, weighted = TRUE)) %>%
  filter(map_lgl(model, ~ inherits(.x, "nls"))) %>%
  mutate(coefficients = map(model, coef)) %>%
  unnest_wider(coefficients)
```

Fits the same double sigmoidal model to each dilution and compares the six coefficients across the dilution series.

## Seed Concentration: Results

```{r Seed vs Parameters}
#| fig-height: 5.5
#| fig-width: 10
seed_mod %>%
  pivot_longer(c(S1, a1, b1, S2, a2, b2), names_to = "parameter", values_to = "value") %>%
  filter(
    value >= (quantile(value, 0.25) - 1.5 * IQR(value)) &
      value <= (quantile(value, 0.75) + 1.5 * IQR(value)),
    .by = parameter
  ) %>%
  mutate(
    hline = ifelse(parameter == "S2", 0, NA),
    parameter = factor(
      parameter,
      levels = c("S1", "a1", "b1", "S2", "a2", "b2"),
      labels = c(TeX("$S_1$"), TeX("$a_1$"), TeX("$b_1$"), TeX("$S_2$"), TeX("$a_2$"), TeX("$b_2$"))
    )
  ) %>%
  ggplot(aes(dilutions, value, color = dilutions)) +
  geom_point(position = position_jitter(width = 0.05)) +
  geom_rug(sides = "l") +
  stat_smooth(method = "loess") +
  geom_line(aes(dilutions, hline), linetype = "dashed", color = "white", linewidth = 1, inherit.aes = FALSE) +
  stat_cor(aes(label = paste(after_stat(rr.label), after_stat(p.label), sep = "~`,`~")), geom = "label", size = 4, fill = "black", color = "white") +
  facet_wrap(vars(parameter), scales = "free_y", labeller = label_parsed) +
  scale_color_gradient(low = "darkcyan", high = "darkorange") +
  labs(x = TeX(r'(Log$_{10}$ Seed Dilution Factor)'), y = "Coefficient Value") +
  main_theme +
  theme(legend.position = "none")
```

- $b_1$ (lag time) shifts with dilution, as expected
- $S_2$ becomes **more negative** as seed concentration increases — more seed, stronger dampening
- $a_1$ and $a_2$ (steepness) are largely unaffected by dilution

## Experiment 2 — Substrate Concentration {.smaller}

**Question:** Does substrate concentration modulate the secondary phase, and does it behave like classical enzyme kinetics?

**Design:** Fixed seed dilution ($10^{-3}$), substrate concentration varied from 0.01 to 0.4 µg/µL, standard reaction conditions.

```{r Load substrate data}
#| include: false
sub_files <- list.files("raw/substrate_effects", pattern = "*.xlsx", full.names = TRUE)
sub_groups <- c("sample", "well", "dilutions", "rxn")

sub_raw <- sub_files %>%
  map_dfr(get_raw) %>%
  clean_names(replace = c("Sample IDs" = "sample")) %>%
  filter(time > 0) %>%
  normalize("time", "rfu", 8, sub_groups, 3) %>%
  mutate(
    across(c(sample, well, rxn), as.factor),
    dilutions = as.integer(dilutions) / 1000
  ) %>%
  na.omit()

sub_df <- sub_raw %>%
  nest(.by = all_of(sub_groups)) %>%
  estimate_params() %>%
  na.omit()

sub_mod <- sub_df %>%
  mutate(model = pmap(., fit_model, single_only = FALSE, weighted = TRUE)) %>%
  filter(map_lgl(model, ~ inherits(.x, "nls"))) %>%
  mutate(coefficients = map(model, coef)) %>%
  unnest_wider(coefficients) %>%
  mutate(v = 1 / b1)
```

If substrate behaves like a classical enzyme system, reaction rate should rise with $[S]$ and saturate at $V_{max}$.

## Substrate Concentration: Results

```{r Substrate vs Parameters}
#| fig-height: 5.5
#| fig-width: 10
sub_mod %>%
  pivot_longer(c(S1, a1, b1, S2, a2, b2), names_to = "parameter", values_to = "value") %>%
  filter(
    value >= (quantile(value, 0.25) - 1.5 * IQR(value)) &
      value <= (quantile(value, 0.75) + 1.5 * IQR(value)),
    .by = parameter
  ) %>%
  mutate(
    hline = ifelse(parameter == "S2", 0, NA),
    parameter = factor(
      parameter,
      levels = c("S1", "a1", "b1", "S2", "a2", "b2"),
      labels = c(TeX("$S_1$"), TeX("$a_1$"), TeX("$b_1$"), TeX("$S_2$"), TeX("$a_2$"), TeX("$b_2$"))
    )
  ) %>%
  ggplot(aes(dilutions, value, color = dilutions)) +
  geom_point(position = position_jitter(width = 0.01)) +
  geom_rug(sides = "l") +
  stat_smooth(method = "loess") +
  geom_line(aes(dilutions, hline), linetype = "dashed", color = "white", linewidth = 1, inherit.aes = FALSE) +
  stat_cor(aes(label = paste(after_stat(rr.label), after_stat(p.label), sep = "~`,`~")), geom = "label", size = 4, fill = "black", color = "white") +
  facet_wrap(vars(parameter), scales = "free_y", labeller = label_parsed) +
  scale_color_gradient(low = "darkcyan", high = "darkorange") +
  labs(x = "[Substrate] (mg/mL)", y = "Coefficient Value") +
  main_theme +
  theme(legend.position = "none")
```

- $S_1$ rises and $S_2$ falls with $[S]$ — more substrate, stronger dampening
- $b_1$ is **non-monotonic**: it peaks around 0.1 mg/mL before settling to an equilibrium near 30 hr
- Not the smooth saturation curve classical enzyme kinetics would predict — there appears to be an optimal $[S]$

## Determining $V_{max}$ and $K_m$

Treating $1/b_1$ as a reaction velocity and fitting $v = \dfrac{V_{max}[S]}{K_m + [S]}$ up to the peak:

```{r Vmax fit}
#| fig-height: 5
#| fig-width: 9
sub_summary <- sub_mod %>%
  summarize(across(c(v), mean), .by = c(sample, dilutions))

vmax <- max(sub_summary$v)
vmax_substrate <- sub_summary[which.max(sub_summary$v), ]$dilutions
Km <- sub_summary[which.max(sub_summary$v <= vmax / 2), ]$dilutions

sub_to_max <- sub_mod %>% filter(dilutions <= vmax_substrate)

nls_mod <- nls(
  v ~ (Vmax * dilutions) / (Km + dilutions),
  data = sub_to_max,
  start = list(Vmax = vmax, Km = Km),
  algorithm = "port"
)

mic_men <- tibble(
  dilutions = seq(0, 0.4, by = 0.01),
  v = (coef(nls_mod)[1] * dilutions) / (coef(nls_mod)[2] + dilutions)
)

sub_mod %>%
  ggplot(aes(x = dilutions, y = v)) +
  geom_point(position = position_jitter(width = 0.01, seed = 157093759), color = "darkorange") +
  geom_line(data = mic_men, aes(x = dilutions, y = v), color = "darkorange", linewidth = 1) +
  scale_x_continuous(breaks = seq(0, 0.4, by = 0.05)) +
  labs(x = "[Substrate] (mg/mL)", y = TeX("V ($h^{-1}$)")) +
  main_theme
```

Theoretical $V_{max} = `r round(coef(nls_mod)[1], 2)`\ h^{-1}$, theoretical $K_m = `r round(coef(nls_mod)[2], 2)`\ mg/mL$.

## Next Steps

- **Tier 1:** Fragmentation dynamics — does shaking speed / temperature shift the secondary phase, implicating mechanical fragmentation?
- **Tier 2:** Orthogonal measurement — do absorbance (A600), SyproOrange, and Bradford assays confirm the secondary phase as a real change in fibril/soluble protein state, or is it ThT-specific?
- **Tier 3:** Generalizability — does the biphasic profile, and its secondary-phase parameters, hold across prion strains and substrate species?

Each tier builds toward the same question: **is the dampening phase biology, or artifact?**

# Thank You

## Learn More

- **Model:** [modeling.gagerowden.com/model.html](https://modeling.gagerowden.com/model.html)
- **Roadmap:** [modeling.gagerowden.com/roadmap.html](https://modeling.gagerowden.com/roadmap.html)
- **Experiments:** [modeling.gagerowden.com/experiments.html](https://modeling.gagerowden.com/experiments.html)

[gagerowden.com](https://gagerowden.com)
