---
title: "Modeling RT-QuIC Kinetics"
date: today
author: "Gage Rowden"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: flatly
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(latex2exp)

main_theme <- theme(
  plot.title = element_text(size = 30, hjust = 0.5),
  axis.title = element_text(size = 24),
  axis.text = element_text(size = 20),
  legend.title = element_text(size = 24),
  legend.text = element_text(size = 20),
  strip.text = element_text(size = 24),
  panel.background = element_rect(fill = "white"),
  panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
)
```

## Introduction
This R package contains code for modeling the kinetics of the enzymes in the RT-QuIC assay. The enzymes are modeled using a two-component model, with the growth rate and decay rate of the enzyme being modeled separately.

The model is described by the following equations:

$$
\begin{aligned}
  \text{primary} &= \frac{P}{1 + e^{a_1(b_1 - \text{time})}} \\
  \text{secondary} &= \frac{S}{1 + e^{a_2(b_2 - \text{time})}}
\end{aligned}
$$

where $P$ and $S$ represent the primary and secondary phase asymptotes, respectively. Coefficients $a_1$ and $a_2$ represent the midpoint of the inflection point of the curves. And, $b_1$ and $b_2$, modulate the steepness of the curves. The model is fit to the data using nonlinear least squares `nls()`.


```{r}
#| fig-height: 12
fit_single <- function(x, a, b, c) {
  return(a / (1 + exp(c * (b - x))))
}

coefs <- tibble(
  P1 = 5, P2 = 10, P3 = 15,
  a1 = 20, a2 = 30, a3 = 40,
  b1 = 1, b2 = 0.5, b3 = 0.25
)

coefs_long <- coefs %>%
  pivot_longer(everything(), names_to = "coefficient", values_to = "value")

combos <- expand.grid(letters = c("P", "a", "b"), numbers = 1:3)

df_coef <- tibble(
  coefficient = paste0(combos$letters, combos$numbers),
  x = c(
    30, coefs$a1 - 5, 30, 
    30, coefs$a2 - 5, 30, 
    30, coefs$a3 - 5, 30),
  y = c(
    coefs$P1 - 1, coefs$P3 / 2, coefs$P3 / 2 + 4, 
    coefs$P2 - 1, coefs$P3 / 2, coefs$P3 / 2, 
    coefs$P3 - 1, coefs$P3 / 2, coefs$P3 / 2 - 4), 
) %>%
  left_join(coefs_long, by = "coefficient") %>%
  mutate(
    coef = str_extract(coefficient, "^[A-z]"),
    mod_number = as.factor(str_extract(coefficient, "[0-9]$")),
    label = paste(coef, "=", value)
  )

df_ <- tibble(
  time = seq(0, 48, 0.25),
  P1 = fit_single(time, coefs$P1, coefs$a1, coefs$b1),
  P2 = fit_single(time, coefs$P2, coefs$a1, coefs$b1),
  P3 = fit_single(time, coefs$P3, coefs$a1, coefs$b1),
  a1 = fit_single(time, coefs$P3, coefs$a1, coefs$b1),
  a2 = fit_single(time, coefs$P3, coefs$a2, coefs$b1),
  a3 = fit_single(time, coefs$P3, coefs$a3, coefs$b1),
  b1 = fit_single(time, coefs$P3, coefs$a1, coefs$b1),
  b2 = fit_single(time, coefs$P3, coefs$a1, coefs$b2),
  b3 = fit_single(time, coefs$P3, coefs$a1, coefs$b3)
) %>%
  pivot_longer(-time, names_to = "parameter", values_to = "value") %>%
  mutate(
      coef = case_when(
        str_detect(parameter, "P") ~ "P", 
        str_detect(parameter, "a") ~ "a", 
        str_detect(parameter, "b") ~ "b"
      ),
      mod_number = as.factor(str_extract(parameter, "[0-9]$"))
  )

df_eq <- df_coef %>%
  summarize(
    .by = coef
  ) %>%
    mutate(
      label = case_when(
        coef == "P" ~ TeX(r"($f(t)=\frac{\textbf{P}}{1+e^{a(b - t)}}$)", output = "character"),
        coef == "a" ~ TeX(r"($f(t)=\frac{P}{1+e^{\textbf{a}(b - t)}}$)", output = "character"),
        coef == "b" ~ TeX(r"($f(t)=\frac{P}{1+e^{a(\textbf{b} - t)}}$)", output = "character")
      )
    )

df_ %>%
  mutate(coef = factor(coef, levels = c("P", "a", "b"))) %>%
  ggplot(aes(time, value, color=mod_number)) +
  geom_line(show.legend = FALSE) +
  geom_text(aes(x=x, y=y, label = label), data = df_coef, size=6, show.legend = FALSE) +
  geom_text(aes(label = label), data = df_eq, x = 0, y = 13, size = 6, hjust = 0, parse = TRUE, inherit.aes = FALSE) +
  facet_grid(rows=vars(fct_inorder(coef)), scales = "free_y") +
  labs(x = "Time (hr)", y = "Normalized Fluorescence", title = "Fit to Single Sigmoid Model") +
  main_theme +
  theme(
    legend.position = "bottom",
    strip.text = element_blank()
  )
```