---
title: Substrate Dilution Series
date: 2026-08-20
---

## Purpose
The $V_{max}$ of our prion seed can be determined by performing a dilution series of the substrate. The rate of the reaction should reach some asymptote as the substrate concentration increases. This experiment will hopefully determine that asymptote for a range of substrate concentrations.

```{r Initialize, echo=FALSE}
library(tidyverse)
library(kableExtra)
library(quicR)
library(janitor)
library(zoo)
library(modelr)
library(ggtext)
library(latex2exp)
library(ggpubr)
library(skimr)

list.files("./R", full.names = TRUE) %>%
  walk(source)
```

## Design

- **Positive seed sample**: 141234
- **Seed dilution factor**: 10^-3^
- **Temperature**: 42°C
- **Shaking speed**: 700rpm
- **Shaking time**: 60s
- **Resting time**: 60s
- **Read cycle time**: 15min
- **Reaction time**: 96 hr
- **Substrate**: HaPrP(90-231)
- **Substrate concentrations**: 0.01 $\rightarrow$ 0.4ug/uL

### Volumes and Concentrations

```{r Plate Design, echo=FALSE}
#| tbl-cap: Volumes and concentrations. Variable amounts of substrate and water will need to be added to each well in the plate. 2uL of seed were added along with 14.6uL of RT-QuIC buffer concentrate for a total reaction volume of 100uL. The starting concentration of the substrate was 0.74ug/uL.
starting_sub_conc <- 0.74
v2 <- 100
buffer_vol <- 14.6
seed_vol <- 2
tibble(
  c2 = signif(seq(0.01, 0.4, length.out = 24), 2),
  v1 = signif(seq(0.01, 0.4, length.out = 24), 2),
  water_vol = signif(seq(0.01, 0.4, length.out = 24), 2),
) %>%
  rename(
    "Substrate Conc (ug/ul)" = 1,
    "Substrate Vol (uL)" = 2,
    "Water (uL)" = 3,
  ) %>%
  kable(align="c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 
```

### Plate Layout of Substrate Concentrations (ug/uL)
|       | **1**  | **2**  | **3**  | **4**  | **5**  | **6**  | **7**  | **8**  | **9**  | **10** | **11** | **12** |
| :---: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: |
| **A** | 0.010  | 0.044  | 0.078  | 0.110  | 0.150  | 0.180  | 0.210  | 0.250  | 0.280  | 0.320  | 0.350  | 0.380  |
| **B** |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |
| **C** |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |
| **D** | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ |
| **E** | 0.027  | 0.061  | 0.095  | 0.130  | 0.160  | 0.200  | 0.230  | 0.260  | 0.300  | 0.330  | 0.370  | 0.400  |
| **F** |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |
| **G** |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |
| **H** | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ |