---
title: Substrate Dilution Series
date: 2026-08-20
---

## Purpose
The $V_{max}$ of our prion seed can be determined by performing a dilution series of the substrate. The rate of the reaction should reach some asymptote as the substrate concentration increases. This experiment will hopefully determine that asymptote for a range of substrate concentrations.

```{r Initialize, echo=FALSE}
library(tidyverse)
library(kableExtra)
library(quicR)
library(janitor)
library(modelr)
library(zoo)
library(ggpubr)
library(magrittr)
library(latex2exp)

list.files("./R", full.names = TRUE) %>%
  walk(source)
```

## Design

- **Positive seed sample**: 141234
- **Seed dilution factor**: 10^-3^
- **Temperature**: 42°C
- **Shaking speed**: 700rpm
- **Shaking time**: 60s
- **Resting time**: 60s
- **Read cycle time**: 15min
- **Reaction time**: 96 hr
- **Substrate**: HaPrP(90-231)
- **Substrate concentrations**: 0.01 $\rightarrow$ 0.4ug/uL

### Volumes and Concentrations

```{r Plate Design, echo=FALSE}
#| tbl-cap: Volumes and concentrations. Variable amounts of substrate and water will need to be added to each well in the plate. 2uL of seed were added along with 14.6uL of RT-QuIC buffer concentrate for a total reaction volume of 100uL. The starting concentration of the substrate was 0.74ug/uL.
starting_sub_conc <- 0.74
v2 <- 100
buffer_vol <- 14.6
seed_vol <- 2
tibble(
  c2 = signif(seq(0.01, 0.4, length.out = 24), 2),
  v1 = signif(seq(0.01, 0.4, length.out = 24), 2),
  water_vol = signif(seq(0.01, 0.4, length.out = 24), 2),
) %>%
  rename(
    "Substrate Conc (ug/ul)" = 1,
    "Substrate Vol (uL)" = 2,
    "Water (uL)" = 3,
  ) %>%
  kable(align="c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 
```

### Plate Layout of Substrate Concentrations (ug/uL)
|       | **1**  | **2**  | **3**  | **4**  | **5**  | **6**  | **7**  | **8**  | **9**  | **10** | **11** | **12** |
| :---: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: | :----: |
| **A** | 0.010  | 0.044  | 0.078  | 0.110  | 0.150  | 0.180  | 0.210  | 0.250  | 0.280  | 0.320  | 0.350  | 0.380  |
| **B** |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |
| **C** |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |
| **D** | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ |
| **E** | 0.027  | 0.061  | 0.095  | 0.130  | 0.160  | 0.200  | 0.230  | 0.260  | 0.300  | 0.330  | 0.370  | 0.400  |
| **F** |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |
| **G** |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |   \|   |
| **H** | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ | $\vee$ |

## Results
```{r Load raw data, echo=FALSE}

files <- list.files("./raw/substrate_effects", pattern = "*.xlsx", full.names = TRUE)
group_list <- c("sample", "well", "dilutions", "rxn")

get_raw <- function(file) {
  file %>%
    get_quic() %>%
    mutate(
      rxn = str_extract(file, "2026\\d{4}_r\\d+_[A-Z]+_\\D+\\."),
      rxn = str_remove(rxn, "\\.")
    )
}

df_raw <- files %>%
  map_dfr(get_raw) %>%
  clean_names(replace = c("Sample IDs" = "sample")) %>%
  filter(time > 0) %>%
  normalize("time", "rfu", 8, group_list, 3) %>%
  mutate(
    across(c(sample, well, rxn), as.factor),
    dilutions = as.integer(dilutions) / 1000,
  ) %>%
  na.omit()

df_ <- df_raw %>%
  nest(.by = all_of(group_list)) %>%
#   left_join(
#     calculate_metrics(
#       df_raw, group_list, threshold = 3,
#       time_col = "time", ttt_values = "norm", auc_values = "norm",
#       norm_col = "norm", deriv_col = "deriv"
#     ),
#     by = group_list
#   ) %>%
#   filter(MPR > 3) %>%
  estimate_params() %>%
  na.omit()
```

```{r Modeling, echo=FALSE}
df_mod <- df_ %>%
  mutate(model = pmap(
    ., fit_model,
    single_only = FALSE, weighted = TRUE, .progress = TRUE
  )) %>%
  filter(map_lgl(model, ~ inherits(.x, "nls"))) %>%
  mutate(
    data = map2(model, data, ~ {
      cc <- coef(.x)
      ci <- tryCatch(
        predFit(.x, newdata = .y, interval = "confidence", level = 0.95),
        error = function(e) NULL
      )
      .y %>%
        add_predictions(.x) %>%
        add_residuals(.x) %>%
        mutate(
          lower = if (is.null(ci)) NA_real_ else ci[, "lwr"],
          upper = if (is.null(ci)) NA_real_ else ci[, "upr"],
          growth = cc[1] / (1 + exp(cc[2] * (cc[3] - time))),
          decay  = cc[4] / (1 + exp(cc[5] * (cc[6] - time))),
        )
    }),
    coefficients = map(model, coef),
    rse = map_dbl(model, sigma),
    aic = map_dbl(model, AIC),
    bic = map_dbl(model, BIC),
    pseudo_r2 = map_dbl(data, ~ cor(.x$pred, .x$norm)^2),
  ) %>%
  unnest_wider(coefficients) %>%
  mutate(
    Kapp = 1 / b1,
    tlagprime = b1 - a1 / 2,
    intercept = (S1 / 2) - a1 * b1,
    # tlag = -intercept / a1,
    tlag = b1 - (S1 / (2 * a1))
  )
```

### Residuals
```{r Residuals, echo=FALSE}
#| fig-height: 12
#| fig-width: 20
df_mod %>%
  unnest(data) %>%
  ggplot(aes(resid)) +
  geom_histogram(bins = 100, position = "identity", fill = "blue") +
  scale_x_continuous(limits = c(-1, 1)) +
  labs(x = "Residuals", y = "Count", title = "Histogram of Residuals") +
  dark_theme +
  theme(
    legend.title = element_blank(),
    legend.position = "inside",
    legend.position.inside = c(0.1, .95),
    legend.justification = c(0, 1),
    legend.background = element_blank(),
    legend.direction = "horizontal",
  )
```

### Effect of substrate concentration on model coefficients
#### S~1,2~
The model asymptotes have both a positive and negative relationship with the substrate concentration ([S]) for S~1~ and S~2~, respectively. The positive relationship with S~1~ is expected given that there is more substrate incorporated into fibrils which would directly result in a higher fluorescent signal.

The negative relationship with S~2~ is interesting. This indicates that the dampening effect increases with [S]. Previously, we observed that the dampening effect had a negative relationship with the rate of the reaction. However, this isn't necessarily the case here because the rate increased only to a point, then decreased as [S] increased.

#### a~1,2~
The apparent rate constant (a~1~) stayed more or less constant through every [S]. I may be wrong, but I believe this is showing the rate constant of our specific seed (141234).

The secondary rate constant (a~2~) had an initial positive relationship with [S], but reached an equilibrium around 0.2 mg/mL. These low initial values could be explained by S~2~ being near zero, and thus their values could be meaningless, and once an accurate S~2~ was established, the a~2~ values resolved around the true a~2~ value.

#### b~1,2~
The lag time (b~1~) behaved as expected up to 0.1 mg/mL [S], at which point it peaked around 20 h and then decreased to an equilibrium around 30 h. The peak effect is unexpected if we are framing this in terms of enzyme kinetics. We would expect a smooth curve approaching $1/V_{max}$, but clearly there is an optimal [S] for this reaction.

The same effect was observed for b~2~, but this is expected since generally b~2~ will only be modeled around the same time as b~1~.

```{r Parameters vs. substrate concentration, echo=FALSE}
#| fig-height: 12
#| fig-width: 20
#| fig-label: fig-coefs
df_mod %>%
  pivot_longer(c(S1, a1, b1, S2, a2, b2), names_to = "parameter", values_to = "value") %>%
  filter(
    value >= (quantile(value, 0.25) - 1.5 * IQR(value)) &
      value <= (quantile(value, 0.75) + 1.5 * IQR(value)),
    .by = parameter
  ) %>%
  mutate(
    hline = ifelse(parameter == "S2", 0, NA),
    parameter = factor(
      parameter, 
      levels = c("S1", "a1", "b1", "S2", "a2", "b2"), 
      labels = c(TeX("$S_1$"), TeX("$a_1$"), TeX("$b_1$"), TeX("$S_2$"), TeX("$a_2$"), TeX("$b_2$"))
    ),
  ) %>%
  ggplot(aes(dilutions, value, color = dilutions)) +
  geom_point(position = position_jitter(width = 0.01)) +
  geom_rug(sides = "l") +
  stat_smooth(method = "loess") +
  geom_line(aes(dilutions, hline), linetype = "dashed", color="white", linewidth = 1, inherit.aes = FALSE) +
  stat_cor(aes(label = paste(after_stat(rr.label), after_stat(p.label), sep = "~`,`~")), geom="label", size = 6, fill = "black", color = "white") +
  facet_wrap(vars(parameter), scales = "free_y", labeller = label_parsed) +
  scale_color_gradient(low = "darkcyan", high = "darkorange") +
  labs(
    title = TeX(
      r'(Predicted Parameters vs. Substrate Concentration Using $f(t)=\frac{\textbf{S_1}}{1+e^{\textbf{a_1}(\textbf{b_1}-t)}}+\frac{\textbf{S_2}}{1+e^{\textbf{a_2}(\textbf{b_2}-t)}}$)'
    ),
    x = "[Substrate] (mg/mL)",
    y = "Coefficient Value",
    color = "Dilution Factor"
  ) +
  dark_theme +
  theme(
    # legend.title = element_blank(),
    legend.position = "inside",
    legend.position.inside = c(0.99, 0.99),
    legend.justification = c(1, 1),
    legend.direction = "horizontal",
    legend.title.position = "top",
    legend.key.width = unit(1, "cm"),
    legend.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "cm"),
    legend.background = element_blank()
  )
# ggsave("figures/kinetics/seed_vs_parameters.png", width = 18, height = 12)
```

### Reaction rate vs. substrate concentration

```{r Vmax vs. substrate concentration, echo=FALSE}
df_mod %<>% mutate(v = 1 / b1)

df_mod_summary <- df_mod %>%
  summarize(
    across(c(v), mean),
    .by = c(sample, dilutions)
  )

vmax <- max(df_mod_summary$v)
vmax_substrate <- df_mod_summary[which.max(df_mod_summary$v), ]$dilutions
Km <- df_mod_summary[which.max(df_mod_summary$v <= vmax / 2), ]$dilutions

df_mod_to_max <- df_mod %>%
  filter(dilutions <= vmax_substrate)

nls_mod <- nls(
  v ~ (Vmax * dilutions) / (Km + dilutions), 
  data = df_mod_to_max, 
  start = list(Vmax = vmax, Km = Km),
  algorithm = "port"
)

v_half <- coef(nls_mod)[1] / 2

df_mic_men <- tibble(
  dilutions = seq(0, 0.4, by = 0.01),
  v = (coef(nls_mod)[1] * dilutions) / (coef(nls_mod)[2] + dilutions)
)
```

Reaction velocity (V) was roughly calculated as the inverse of the b~1~ coefficient. These values were plotted against the substrate concentration and a non-linear regression was fitted to the data up to the peak [S] mentioned above. This gave us a theoretical V~max~ of `r round(coef(nls_mod)[1], 2)` h^-1^ and a theoretical K~m~ of `r round(coef(nls_mod)[2], 2)` mg/mL.

```{r Vmax vs. [S] plot, echo=FALSE}
#| fig-height: 6
#| fig-width: 10

df_mod %>%
  # filter(dilutions <= vmax_substrate) %>%
  ggplot(aes(x = dilutions, y = v)) +
  geom_point(position = position_jitter(width = 0.01, seed = 157093759), color = "darkorange") +
  geom_line(data = df_mic_men, aes(x = dilutions, y = v), color = "darkorange", linewidth = 1) +
  geom_segment(aes(x = coef(nls_mod)[2], xend = coef(nls_mod)[2], y = 0, yend = v_half), color = "darkorange", linewidth = 1, linetype = "dashed") +
  geom_text(
    aes(label = sprintf(
      "Theoretical Vmax = %s\nTheoretical Km = %s", 
      round(coef(nls_mod)[1], 2), 
      round(coef(nls_mod)[2], 2)
    )), x = 0, y = 0.25, color = "darkorange", size = 6, hjust = 0) +
  # stat_smooth(method = "loess", color = "darkorange", fullrange = TRUE, method.args = list(degree = 2)) +
  scale_x_continuous(breaks = seq(0, 0.4, by = 0.05)) +
  labs(
    x = "[Substrate] (mg/mL)",
    y = TeX("V ($h^{-1}$)")
  ) +
  dark_theme
```